---
title: "Read in phenotype information for Framingham data"
author: 
  - David lukacsovich^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    highlight: pygments
    theme: yeti
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

# Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r library}
suppressMessages({
  library(dplyr)                                                                   # Data manipulation (filter, mutate, select, joins)
  library(tidyr)                                                                   # Reshaping data (pivot_longer, replace_na)
  library(purrr)                                                                   # Functional mapping over file lists (map)
  library(data.table)                                                              # Fast I/O (fread) for large phenotype/sample files
  library(stringr)                                                                 # String utilities (str_sub, str_remove)
})

base_dir    <- "../../.."                                                          # Project root (relative to script location)
analysis_dir <- file.path(base_dir, "DATASET", "Framingham")                       # Root directory for Framingham dataset bundle
raw_dir     <- file.path(analysis_dir, "DNAm/raw/EPIC")                            # Location of raw EPIC methylation assets
pheno_dir   <- file.path(analysis_dir, "Phenotype/raw")                            # Location of raw phenotype releases
data_dir    <- file.path(analysis_dir, "DNAm/processed/EPIC")                      # Location of processed DNAm outputs (not used here, kept for parity)
harmony_dir <- file.path(base_dir, "DATASET/_HARMONIZATION/data_CN_only")          # Harmonization workspace (CN-only context)
save_dir    <- file.path(harmony_dir, "preprocessing/FHS")                         # Destination folder for saved phenotype file
```

# Get Sample Data

## Read Sample Attributes (from IDAT attribute files)    

```{r read_sample}
pheno_df <- list.files(                                                            # Start: enumerate all matching sample attribute files
  path       = file.path(raw_dir, "sample_info"),                                  # Directory holding sample_info exports
  pattern    = "idat.*.gz",                                                        # Pattern for IDAT attribute tables (gzipped)
  full.names = TRUE                                                                # Return full file paths for fread
) %>%                                                                              # Pipe file list into map/read/bind pipeline
  purrr::map(., ~data.table::fread(., skip = 10)) %>%                              # Read each file, skipping dbGaP header lines
  Reduce(rbind, .) %>%                                                             # Row-bind all tables into one data.frame
  dplyr::mutate(                                                                   # Create derived identifiers
    idat_label = stringr::str_sub(.data$Grn.idat, 1, 24)                           # Extract stable IDAT label from file name field
  ) %>%                                                                            # Continue transformation pipeline
  dplyr::select("dbGaP_Sample_ID", "SAMPLE_ID", "Batch", "idat_label") %>%         # Keep only fields needed downstream
  dplyr::mutate(sample = .data$SAMPLE_ID)                                          # Create 'sample' alias (convenience/compatibility)

```

## Get Subject Information

```{r read_subject}
sample_df <- data.table::fread(                                                     # Read sample-to-subject mapping file(s)
  list.files(                                                                       # Enumerate all Sample.MULTI files
    path       = file.path(raw_dir, "sample_info"),                                 # Directory holding sample_info exports
    pattern    = "Sample.MULTI.*.gz",                                               # Pattern for sample mapping tables
    full.names = TRUE                                                               # Return full paths
  ),                                                                                # End list.files for Sample.MULTI
  skip = 10                                                                         # Skip dbGaP header lines
) %>%                                                                               # Pipe into dplyr select
  dplyr::select("dbGaP_Subject_ID", "SAMPLE_ID")                                    # Keep only subject and sample IDs

share_df <- list.files(                                                             # Start: enumerate the Info tables with shareid/date fields
  path       = file.path(pheno_dir, "Info/"),                                       # Phenotype Info folder
  pattern    = "txt.gz",                                                            # Gzipped tab-delimited phenotype files
  full.names = TRUE                                                                 # Return full paths
) %>%                                                                               # Pipe file list into map/read/bind pipeline
  purrr::map(., ~data.table::fread(., skip = 10)) %>%                               # Read each file, skipping dbGaP headers
  Reduce(rbind, .) %>%                                                              # Row-bind into one table
  dplyr::select("dbGaP_Subject_ID", "shareid", "idtype", "date2", "date9")          # Keep IDs and exam date anchors

pheno_df <- pheno_df %>%                                                            # Start merging subject-level identifiers onto sample table
  dplyr::left_join(sample_df, by = "SAMPLE_ID") %>%                                 # Add dbGaP_Subject_ID using SAMPLE_ID key
  dplyr::left_join(share_df,  by = "dbGaP_Subject_ID")                              # Add shareid/idtype/date2/date9 using subject key
```

## Get Age and Collection Information

```{r read_collection}
age_df <- list.files(                                                               # Enumerate methylation sample attribute tables with age/visit
  path       = file.path(raw_dir, "sample_info"),                                   # Directory holding sample_info exports
  pattern    = "FHS_Methylation_Sample_Attributes.*txt.gz",                         # Pattern for collection-age attribute tables
  full.names = TRUE                                                                 # Return full paths
) %>%                                                                               # Pipe file list into map/read/bind pipeline
  purrr::map(., ~data.table::fread(., skip = 10)) %>%                               # Read each file, skipping dbGaP headers
  Reduce(rbind, .) %>%                                                              # Row-bind into one table
  dplyr::select("SAMPLE_ID", Age = "Age_at_collection", "Collection_visit")         # Keep sample ID, age, and collection visit label

pheno_df <- pheno_df %>%                                                            # Merge age/collection visit onto main phenotype table
  dplyr::left_join(age_df, by = "SAMPLE_ID")                                        # Join by SAMPLE_ID so each sample has age/visit
```

## Calculate time since exam

```{r read_exam_time}
time_df <- pheno_df %>%                                                            # Start from main table containing Collection_visit and dates
  dplyr::select("dbGaP_Sample_ID", "Collection_visit", "date2", "date9") %>%       # Keep only fields needed to compute time since exam
  tidyr::pivot_longer(                                                             # Reshape wide date2/date9 into long format for matching
    cols      = c("date2", "date9"),                                               # Date anchor columns to stack
    names_to  = "date",                                                            # Name of the new column holding which date variable
    values_to = "time_since_exam1"                                                 # Name of the new column holding the date value
  ) %>%                                                                            # Continue transformation pipeline
  dplyr::mutate(                                                                   # Harmonize visit label to match pivoted date variable names
    Collection_visit = paste0("date", .data$Collection_visit)                      # Convert visit number (e.g., 2/9) into "date2"/"date9"
  ) %>%                                                                            # Continue transformation pipeline
  dplyr::filter(.data$Collection_visit == .data$date) %>%                          # Keep only the row matching the sample’s collection visit
  dplyr::select("dbGaP_Sample_ID", "time_since_exam1")                             # Output only sample ID and its matched anchor date

pheno_df <- pheno_df %>%                                                           # Merge computed anchor-date value back into main table
  dplyr::left_join(time_df, by = "dbGaP_Sample_ID") %>%                            # Join by sample ID (dbGaP sample identifier)
  tidyr::replace_na(list(time_since_exam1 = 0))                                    # If missing (no match), set to 0 as a conservative default

```

# Get Sample Attributes

## Get Sex Information

```{r get_sex}
sex_df <- data.table::fread(                                                        # Read subject-level table containing SEX coding
  list.files(                                                                       # Enumerate all Subject.MULTI files
    path       = file.path(raw_dir, "sample_info"),                                 # Directory holding sample_info exports
    pattern    = "Subject.MULTI.*.gz",                                              # Pattern for subject tables
    full.names = TRUE                                                               # Return full paths
  ),                                                                                # End list.files for Subject.MULTI
  skip = 10                                                                         # Skip dbGaP header lines
) %>%                                                                               # Pipe into mutate/select
  dplyr::mutate(Sex = ifelse(.data$SEX == 1, "MALE", "FEMALE")) %>%                 # Recode SEX numeric to labeled Sex factor
  dplyr::select("dbGaP_Subject_ID", "SUBJECT_ID", "Sex")                            # Keep IDs and recoded Sex

pheno_df <- pheno_df %>%                                                            # Merge sex information onto sample-level table
  dplyr::left_join(sex_df, by = "dbGaP_Subject_ID")                                 # Join via dbGaP_Subject_ID to attach Sex
```

## Get APOE Information

```{r get_apoe}
apoe_df <- list.files(                                                              # Enumerate APOE phenotype files
  path       = file.path(pheno_dir, "APOE"),                                        # APOE folder within raw phenotype directory
  pattern    = "csv.gz",                                                            # Gzipped CSV files
  full.names = TRUE,                                                                # Return full paths
  recursive  = TRUE                                                                 # Search subfolders if present
) %>%                                                                               # Pipe file list into map/read/bind pipeline
  purrr::map(., ~data.table::fread(.)) %>%                                          # Read each APOE table (no header skip assumed here)
  Reduce(rbind, .) %>%                                                              # Row-bind into one APOE table
  dplyr::mutate(                                                                    # Create APOE4 allele dosage using coding rules
    APOE4 = ifelse(                                                                 # Start nested ifelse to translate APOE coding to allele count
      APOE == 44,                                                                   # If genotype code indicates ε4/ε4
      2,                                                                            # Then APOE4 dosage is 2
      ifelse(APOE %in% c(24, 34), 1, 0)                                             # Else if ε2/ε4 or ε3/ε4 then 1; otherwise 0
    )                                                                               # End nested ifelse
  ) %>%                                                                             # Continue transformation pipeline
  dplyr::select("shareid", "idtype", "APOE", "APOE4")                               # Keep identifiers and APOE/APOE4 variables

pheno_df <- pheno_df %>%                                                            # Merge APOE fields onto main phenotype table
  dplyr::left_join(apoe_df, by = c("shareid", "idtype"))                            # Join on shareid/idtype (phenotype-system identifiers)
```

## Get Smoking Status Information

```{r get_smoking}
smoking_df <- list.files(                                                           # Enumerate smoking phenotype files
  path       = file.path(pheno_dir, "Smoking/"),                                    # Smoking folder
  pattern    = "txt.gz",                                                            # Gzipped tab-delimited files
  full.names = TRUE                                                                 # Return full paths
) %>%                                                                               # Pipe into map/read/bind pipeline
  purrr::map(., ~data.table::fread(., skip = 10)) %>%                               # Read each file, skipping dbGaP headers
  Reduce(rbind, .) %>%                                                              # Row-bind into one smoking table
  dplyr::rename(idtype = "IDTYPE") %>%                                              # Standardize IDTYPE column name to idtype for joins
  tidyr::pivot_longer(                                                              # Convert STATUS1:STATUS9 columns into long format
    cols      = STATUS1:STATUS9,                                                    # Smoking status fields across exam visits
    names_to  = "Status",                                                           # Store original STATUS column name
    values_to = "Smoking"                                                           # Store smoking value in a single column
  ) %>%                                                                             # Continue transformation pipeline
  dplyr::mutate(                                                                    # Parse visit number from STATUS column name
    Collection_visit = as.integer(stringr::str_remove(.data$Status, "STATUS"))      # Convert "STATUS9" -> 9, etc.
  ) %>%                                                                             # Continue transformation pipeline
  dplyr::select(-"Status")                                                          # Drop temporary Status column

pheno_df <- pheno_df %>%                                                            # Merge visit-matched smoking status onto main table
  dplyr::left_join(                                                                 # Left join so samples remain even if smoking missing
    smoking_df,                                                                     # Long-format smoking status table
    by = c("dbGaP_Subject_ID", "shareid", "idtype", "Collection_visit")             # Join on subject IDs and collection visit
  )                                                                                 # End left_join
```

## Get Education Status

```{r get_education}
edu_df <- list.files(                                                               # Enumerate education phenotype files
  path       = file.path(pheno_dir, "Education/"),                                  # Education folder
  pattern    = "txt.gz",                                                            # Gzipped tab-delimited files
  full.names = TRUE                                                                 # Return full paths
) %>%                                                                               # Pipe into map/read/bind pipeline
  purrr::map(., ~data.table::fread(., skip = 10)) %>%                               # Read each file, skipping dbGaP headers
  Reduce(rbind, .) %>%                                                              # Row-bind into one education table
  dplyr::rename(Education_level = "education")                                      # Rename to explicit Education_level variable

exam2_df <- list.files(                                                             # Enumerate Exam2 phenotype files (for education years)
  path       = file.path(pheno_dir, "Exam2/"),                                      # Exam2 folder
  pattern    = "txt.gz",                                                            # Gzipped tab-delimited files
  full.names = TRUE                                                                 # Return full paths
) %>%                                                                               # Pipe into map/read/bind pipeline
  purrr::map(., ~data.table::fread(., skip = 10)) %>%                               # Read each file, skipping dbGaP headers
  Reduce(rbind, .) %>%                                                              # Row-bind into one Exam2 table
  dplyr::rename(Education_year = "B43") %>%                                         # Rename B43 to Education_year (years of education)
  dplyr::select("dbGaP_Subject_ID", "shareid", "idtype", "Education_year")          # Keep only fields needed for merge

pheno_df <- pheno_df %>%                                                            # Merge both education indicators onto main table
  dplyr::left_join(edu_df,   by = c("dbGaP_Subject_ID", "shareid", "idtype")) %>%   # Attach Education_level by IDs
  dplyr::left_join(exam2_df, by = c("dbGaP_Subject_ID", "shareid", "idtype"))       # Attach Education_year by IDs
```

## Get Dementia Status

```{r get_dementia}
dem_df <- list.files(                                                               # Enumerate dementia survival phenotype files
  path       = file.path(pheno_dir, "Dementia/"),                                   # Dementia folder
  pattern    = "demsurv.*txt.gz",                                                   # Pattern for dementia survival tables
  full.names = TRUE                                                                 # Return full paths
) %>%                                                                               # Pipe into map/read/bind pipeline
  purrr::map(., ~data.table::fread(., skip = 10)) %>%                               # Read each file, skipping dbGaP headers
  Reduce(rbind, .) %>%                                                              # Row-bind into one dementia survival table
  dplyr::select(                                                                    # Keep only fields needed for merge
    "dbGaP_Subject_ID",                                                             # Keep dbGaP_Subject_ID
    "shareid",                                                                      # Keep shareid
    "idtype",                                                                       # Keep idtype
    "DEM_SURVDATE",                                                                 # Keep number of days of survival after Dementia
    Dementia = "DEM_STATUS"                                                         # Keep dementia status
  )                                                                                 # End select block

pheno_df <- pheno_df %>%                                                            # Merge dementia fields and derive follow-up times
  dplyr::left_join(dem_df, by = c("dbGaP_Subject_ID", "shareid", "idtype")) %>%     # Join dementia outcomes by phenotype-system IDs
  dplyr::mutate(                                                                    # Compute follow-up relative to baseline exam anchor
    dem_followup_time = .data$DEM_SURVDATE - .data$time_since_exam1,                # Time from baseline exam anchor to event/censoring date
    dem_followup_year = round(.data$dem_followup_time / 365)                        # Approximate follow-up in years (rounded)
  )                                                                                 # End mutate block
```

# Filter to Target Data

## Filter to Offspring Cohort

```{r filter_cohort}
pheno_df <- pheno_df %>%                                                            # Start filtering pipeline
  dplyr::filter(.data$idtype == 1)                                                  # Keep only offspring cohort records
```

## Filter by Dementia Status

```{r filter_dementia}
pheno_df <- pheno_df %>%
pheno_df <- pheno_df %>%                                                            # Start filtering pipeline
  dplyr::filter(.data$Dementia == 0) %>%                                            # Keep only dementia-negative at baseline
  dplyr::filter(.data$dem_followup_time > 0)                                        # Keep only those with positive follow-up time
```

# Save

```{r save}
write.csv(                                                                         # Write final phenotype table to CSV
  pheno_df,                                                                        # Data frame to export
  file = file.path(save_dir, "phenotype_all.csv"),                                 # Output path (harmonization preprocessing folder)
  row.names = FALSE                                                                # Do not write row names into CSV
)               
```

# Session Information

<details>
  <summary>**Session Info**</summary>
```{r session}
sessionInfo()                                                                   # Print session/package versions for reproducibility
```
</details>
