---
title: "Read in HRS NIAGADS data, get detection P scores, filter out unwanted probes, and convert to beta values"
author: 
  - David lukacsovich^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    highlight: pygments
    theme: yeti
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up

## Load Libraries

```{r library}
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(minfi)
  library(DMRcate)
  library(SummarizedExperiment)
})

base_dir <- "~/TBL Dropbox/David Lukacsovich"
hrs_dir <- file.path(base_dir, "DATASET", "HRS")
ref_dir <- file.path(hrs_dir, "DNAm", "ref_files")
code_dir <- file.path(hrs_dir, "R")

analysis_dir <- file.path(base_dir, "DATASET", "MESA")
data_dir <- file.path(analysis_dir, "DNAm/processed/EPIC") 

source(file.path(code_dir, "run_parallel.R"), local = TRUE)
source(file.path(code_dir, "detectionp_functions.R"), local = TRUE)
```

## Read Data

```{r load_data}
dnam_data <- readRDS(file.path(data_dir, "DNAm_Data.RDS"))
```

# detectionP

## Get detectionP scores

```{r detection_p}
assay_data <- SummarizedExperiment::assays(dnam_data)$dnam
detp_df <- data.frame(
  sample = character(),
  cpg = character()
)

for (plate in unique(dnam_data$batch)) {
  print(paste0("detectionP on : ", plate))
  detp_df <- rbind(
    detp_df,
    detectionp_plate(dnam_data, plate, threshold = 0.01, plate_col = "batch")
  )
}
```

## Get Probe Annotation

```{r get_annot}
anno_df <- read.csv(
  file.path(ref_dir, "infinium-methylationepic-v-1-0-b5-manifest-file.csv"),
  skip = 7
) %>%
  dplyr::select(
    "Name",
    seqnames = "CHR_hg38",
    pos = "Start_hg38",
    Relation_to_Island = "Relation_to_UCSC_CpG_Island",
    "UCSC_RefGene_Name",
    "UCSC_RefGene_Group",
    type12 = "Infinium_Design_Type"
  ) %>%
  dplyr::filter(
    (!is.na(.data$pos)) & (nchar(.data$type12) > 0)
  ) %>%
  dplyr::filter(.data$seqnames != "chrM")

row.names(anno_df) <- anno_df$Name
```

## Convert to beta values

```{r convert_beta}
assay_data <- minfi::getBeta(assay_data)
assay_data <- assay_data[row.names(assay_data) %in% row.names(anno_df),]
anno_df <- anno_df[row.names(assay_data),]

pheno_df <- dnam_data %>%
  SummarizedExperiment::colData() %>%
  as.data.frame()
dnam_data <- SummarizedExperiment::SummarizedExperiment(
  assays = list(dnam = assay_data),
  rowData = anno_df,
  colData = pheno_df
)
```

# Probe Filtering

## Filter to Probes Starting with 'cg'

```{r filter_cg}
dnam_data <- dnam_data[grep("cg", row.names(assay_data)), ]
```

## Filter Probes by SNP distance, cross hybridization

```{r filter_snp}
assay_data <- SummarizedExperiment::assays(dnam_data)$dnam

assay_data <- DMRcate::rmSNPandCH(
  assay_data,
  dist = 5,
  mafcut = 0.01,
  and = TRUE,
  rmcrosshyb = TRUE,
  rmXY = FALSE
)

row_df <- dnam_data %>%
  SummarizedExperiment::rowData() %>%
  as.data.frame() %>%
  dplyr::mutate(filter_snp = !(.data$Name %in% row.names(assay_data))) %>%
  dplyr::mutate(filter_xy = .data$seqnames %in% c("chrX", "chrY"))
```

## Relabel Rows

```{r se_final}
assay_data <- SummarizedExperiment::assays(dnam_data)$dnam
pheno_df <- dnam_data %>%
  SummarizedExperiment::colData() %>%
  as.data.frame()
dnam_data <- SummarizedExperiment::SummarizedExperiment(
  assays = list(dnam = assay_data),
  rowData = row_df,
  colData = pheno_df
)
```

## Filter detection P probes

Some of the probes are missing from the probe annotation file. We filter those out, so they also need to be filtered out from the detectionP results

```{r filter_detp}
detp_df <- detp_df %>%
  dplyr::filter(.data$cpg %in% row.names(dnam_data))
```

# Get NA Probes

## Filter to Probes and Samples with NAs

```{r adjust_na}
na_data <- is.na(assay_data)
row_counts <- rowSums(na_data)
na_data <- na_data[row_counts > 0,]
col_counts <- colSums(na_data)
na_data <- na_data[,col_counts > 0]
```

## Get NA dataframe

```{r data_na}
na_df <- which(na_data, arr.ind = TRUE) %>%
  as.data.frame()

inds <- cbind(na_df$row, na_df$col)

cpg_df <- data.frame(
  row = 1:nrow(na_data),
  cpg = row.names(na_data)
)

sample_df <- data.frame(
  col = 1:ncol(na_data),
  sample = colnames(na_data)
)

na_df <- na_df %>%
  dplyr::left_join(cpg_df, by = "row") %>%
  dplyr::left_join(sample_df, by = "col") %>%
  dplyr::select("sample", "cpg")
```

# Filter High Missing Samples

We filter out any samples that have at least 5% of their non-XY probes with high detection P values

## Get Counts

```{r sample_filter_detp}
base_probes <- row_df %>%
  dplyr::filter(!.data$filter_xy) %>%
  dplyr::pull("Name")

base_df <- detp_df %>%
  dplyr::filter(.data$cpg %in% base_probes) %>%
  dplyr::group_by(.data$sample) %>%
  dplyr::summarise(missing_count = n()) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    to_filter = .data$missing_count >= 0.05 * length(base_probes)
  )

to_drop <- base_df %>%
  dplyr::filter(.data$to_filter) %>%
  dplyr::pull("sample")

dnam_data <- dnam_data[,!(dnam_data$sample %in% to_drop)]
na_df <- na_df %>%
  dplyr::filter(!(.data$sample %in% to_drop))
detp_df <- detp_df %>%
  dplyr::filter(!(.data$sample %in% to_drop))
```

# Save Results

```{r save}
saveRDS(dnam_data, file = file.path(data_dir, "Preprocessed_Beta.RDS"))
write.csv(
  detp_df,
  file = file.path(data_dir, "detectionP_Failures.csv"),
  row.names = FALSE
)
write.csv(
  na_df,
  file = file.path(data_dir, "missing_probes.csv"),
  row.names = FALSE
)
row_df <- dnam_data %>%
  SummarizedExperiment::rowData() %>%
  as.data.frame()
pheno_df <- dnam_data %>%
  SummarizedExperiment::colData() %>%
  as.data.frame()
write.csv(
  row_df,
  file = file.path(data_dir, "Preprocssed_Probes.csv"),
  row.names = TRUE
)
write.csv(
  pheno_df,
  file = file.path(data_dir, "Preprocssed_Metadata.csv"),
  row.names = TRUE
)
```

# Session Information

<details>
  <summary>**Session Info**</summary>
```{r session}
sessionInfo()
```
</details>
