---
title: "Evaluate batch-correction using Harman method"
subtitle: "Using Kruskal Wallis test, checking for keeping sex-effects, and keeping Age-effects"
author: 
  - David lukacsovich^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    highlight: pygments
    theme: yeti
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up

## Load Libraries

```{r library}
suppressMessages({
  library(dplyr)
  library(ggplot2)
  library(ggpubr)
  library(gmodels)
  library(SummarizedExperiment)
})


base_dir <- "~/TBL Dropbox/David Lukacsovich"
analysis_dir <- file.path(base_dir, "DATASET", "HRS")
data_dir <- file.path(analysis_dir, "datasets")
norm_dir <- file.path(data_dir, "normalized")
batch_dir <- file.path(data_dir, "batch_corrected")
```

# Define Functions

## Load Data

```{r}
prepare_dnam_metadata <- function(dnam_data, dataset) {
  if (dataset == "ADNI") {
    dnam_data$age_split_by_tertiles <- ntile(dnam_data$age_at_visit, 3)
  } else if (dataset == "AIBL") {
    dnam_data$age_split_by_tertiles <- ntile(dnam_data$pred_age_elastic, 3)
    disease <- dnam_data$disease
    diagnosis <- ifelse(
      disease == "Alzheimer's disease", "AD",
      ifelse(
        disease == "Mild Cognitive Impairment", "MCI",
        "CN"
      )
    )
    dnam_data$diagnosis <- diagnosis
  } else if (dataset == "FHS") {
    dnam_data$age_split_by_tertiles <- ntile(dnam_data$Age_at_collection, 3)
    batches <- dnam_data$batch
    batches <- ifelse(batches == "B0019", "B0018", batches)
    dnam_data$batch <- batches
  } else if (dataset == "MESA") {
    dnam_data$race <- as.character(dnam_data$race1c)
  }
  
  dnam_data
}

load_normalized_data <- function(dataset) {
  dnam_data <- readRDS(
    file.path(norm_dir, paste0(dataset, "_Autosomal.RDS"))
  )
  
  dnam_data <- prepare_dnam_metadata(dnam_data, dataset)
  
  dnam_data
}

load_batched_data <- function(dataset) {
  dnam_data <- readRDS(
    file.path(batch_dir, paste0(dataset, "_Batched.RDS"))
  )
  
  dnam_data
}
```

## PCA

```{r}
OrderDataBySd <- function(exp_mat){
  # compute sds for each row
  sds <- matrixStats::rowSds(exp_mat)
  sdsSorted <- order(sds, decreasing = TRUE)
  
  # order by most variable probes on top
  exp_mat[sdsSorted ,]
}

calculate_dnam_pca <- function(dnam_data) {
  beta_data <- SummarizedExperiment::assays(dnam_data)$dnam
  beta_data <- OrderDataBySd(beta_data)
  pc_vals <- stats::prcomp(
    t(beta_data[1:50000,]),
    center = TRUE,
    scale = TRUE
  )
  
  pc_vals
}

get_age_quantiles <- function(ages) {
  cutoffs <- as.numeric(quantile(ages, c(0.25, 0.5, 0.75)))
  
  conditions <- rep("4th_quant", length(ages))
  conditions[ages <= cutoffs[[3]]] <- "3rd_quant"
  conditions[ages <= cutoffs[[2]]] <- "2nd_quant"
  conditions[ages <= cutoffs[[1]]] <- "1st_quant"
  
  conditions
}

get_dnam_pca <- function(dnam_data) {
  pc_vals <- calculate_dnam_pca(dnam_data)
  pc_df <- data.frame(
    sample_ind = paste0("s", 1:nrow(pc_vals$x)),
    PC_1 = pc_vals$x[,1],
    PC_2 = pc_vals$x[,2],
    PC_3 = pc_vals$x[,3],
    PC_4 = pc_vals$x[,4],
    PC_5 = pc_vals$x[,5]
  )
  pc_df <- cbind(
    pc_df,
    dnam_data %>%
      SummarizedExperiment::colData() %>%
      as.data.frame()
  )
  
  percentVar <- pc_vals$sdev^2 / sum(pc_vals$sdev^2)
  
  list(pc_df = pc_df, var_ls = percentVar[1:5])
}

quantile_pca <- function(pc_df) {
  x_thresh <- pc_df %>%
    dplyr::filter(.data$Status == "Before Batch Correction") %>%
    dplyr::pull("PC_1") %>%
    as.numeric() %>%
    median()
  
  y_thresh <- pc_df %>%
    dplyr::filter(.data$Status == "Before Batch Correction") %>%
    dplyr::pull("PC_2") %>%
    as.numeric() %>%
    median()
  
  quant_df <- pc_df %>%
    dplyr::filter(.data$Status == "Before Batch Correction") %>%
    dplyr::mutate(quadrant = "I") %>%
    dplyr::mutate(
      quadrant = ifelse(
        .data$PC_1 > x_thresh,
        ifelse(.data$PC_2 > y_thresh, "I", "II"),
        ifelse(.data$PC_2 > y_thresh, "IV", "III")
        )
    ) %>%
    dplyr::select("sample_ind", "quadrant")
  
  pc_df <- pc_df %>%
    dplyr::left_join(quant_df, by = "sample_ind")
  
  pc_df
}

get_both_pca <- function(norm_data, batch_data) {
  norm_res <- get_dnam_pca(norm_data)
  batch_res <- get_dnam_pca(batch_data)
  
  pc_df <- rbind(
    norm_res$pc_df %>% dplyr::mutate(Status = "Before Batch Correction"),
    batch_res$pc_df %>% dplyr::mutate(Status = "After Batch Correction")
  ) %>%
    quantile_pca()
  
  var_df <- data.frame(
    pc = paste0("PC", 1:5),
    before = as.numeric(norm_res$var_ls),
    after = as.numeric(batch_res$var_ls)
  )
  
  list(pc_df = pc_df, var_df = var_df)
}

plot_pca <- function(pc_df, percentVar, title, group_col) {
  plot_df <- pc_df
  plot_df[,group_col] <- as.character(pc_df[,group_col])
  
  ggplot2::ggplot(
    plot_df, mapping = ggplot2::aes(x = PC_1, y = PC_2, color = !!sym(group_col))
  ) + 
    ggplot2::geom_point() + 
    ggplot2::theme_bw() + 
    ggplot2::xlab(paste0("PC1: ",round(percentVar[1] * 100),"% variance")) + 
    ggplot2::ylab(paste0("PC2: ",round(percentVar[2] * 100),"% variance")) + 
    ggplot2::ggtitle(title)
}

plot_pair_pca <- function(pc_df, var_df, group_col) {
  p1 <- plot_pca(
    pc_df %>% dplyr::filter(.data$Status == "Before Batch Correction"),
    var_df$before,
    "Before Correction",
    group_col
  )
  
  p2 <- plot_pca(
    pc_df %>% dplyr::filter(.data$Status == "After Batch Correction"),
    var_df$after,
    "After Correction",
    group_col
  )
  
  list(p1 = p1, p2 = p2)
}
```

## KW test

```{r}
create_kw_plot <- function(pc_df, pc) {
  # Perform Kruskal-Wallis Test
  kw_test_results <- pc_df %>%
    dplyr::group_by(.data$Status) %>% 
    rstatix::kruskal_test(data = ., as.formula(paste0(pc, " ~ batch")))
  
  # Merge p-values back to the data for plotting
  
  ## Determine maximum y value for text positioning
  max_y <- pc_df %>%
    dplyr::pull(as.name(pc)) %>%
    max(na.rm = TRUE)
  xpos <- length(unique(pc_df$batch)) / 2
  
  ## Plot
  ggpubr::ggboxplot(
    data = pc_df,
    x = "batch",
    y = pc,
    fill = "Status",
    title = paste0("Boxplot for ", pc)
  ) +
    # Add KW Test Results
    ggplot2::annotate(
      "text",
      label = paste0(
        "KW test p before batch correction = ", 
        signif(
          kw_test_results$p[kw_test_results$Status == "Before Batch Correction"],
          3
        ),
        "\nKW test p after batch correction = ",
        signif(
          kw_test_results$p[kw_test_results$Status == "After Batch Correction"],
          3
        )
      ),
      x = xpos,
      y = max_y + 1, 
      size = 4
    ) +
    ggplot2::theme(axis.text.x = ggplot2::element_blank())
}

create_kw_plotlist <- function(pc_df, pc_ls) {
  plyr::llply(
    pc_ls,
    .fun = function(pc) {
      create_kw_plot(pc_df, pc)
    }
  )
}
```

## Check Confounding Effects

```{r}
check_confounding <- function(dnam_data, expt_var) {
  pheno_tab <- dnam_data %>%
    SummarizedExperiment::colData() %>%
    as.data.frame() %>%
    dplyr::select("batch", as.name(expt_var)) %>%
    table()
  
  pheno_sum <- pheno_tab %>%
    summary()
  
  list(tab = pheno_tab, summ = pheno_sum)
}
```

## Wrapper Function

```{r}
analyze_dataset <- function(dataset, expt_var) {
  # load dnam data
  norm_data <- load_normalized_data(dataset)
  batch_data <- load_batched_data(dataset)
  
  # calculate pca parameters
  pca_res <- get_both_pca(norm_data, batch_data)
  pc_df <- pca_res$pc_df
  var_df <- pca_res$var_df
  
  # create pca plots
  pca1 <- plot_pair_pca(pc_df, var_df, "batch")
  pca2 <- plot_pair_pca(pc_df, var_df, expt_var)
  pca3 <- plot_pair_pca(pc_df, var_df, "quadrant")
  
  # create kw plots
  kw_ls <- create_kw_plotlist(pc_df, paste0("PC_", 1:5))
  
  # return results
  list(
    pca1 = pca1,
    pca2 = pca2,
    pca3 = pca3,
    kw_ls = kw_ls
  )
}

display_confounding <- function(dnam_data, expt_var, label) {
  message(paste0("Evaluating Confounding by: ", label))
  
  conf_res <- check_confounding(dnam_data, expt_var)
  conf_tab <- conf_res$tab
  conf_summ <- conf_res$summ
  message("Table")
  print(conf_tab)
  message("Summary")
  print(conf_summ)
  message("")
}

display_age_corr <- function(dnam_data, age_var) {
  pheno_df <- dnam_data %>%
    SummarizedExperiment::colData() %>%
    as.data.frame() %>%
    dplyr::select("batch", as.name(age_var))
  
  kw_test_results <- pheno_df %>%
    rstatix::kruskal_test(data = ., as.formula(paste0(age_var, " ~ batch"))) %>%
    as.data.frame()
  
  message("Evaluating Correlation with Age")
  print(kw_test_results)
}

analyze_confounding <- function(dataset) {
  norm_data <- load_normalized_data(dataset)
  
  if (dataset == "ADNI") {
    display_confounding(norm_data, "age_split", "Age Tertiles")
    display_confounding(norm_data, "sex", "Sex")
    display_confounding(norm_data, "DX", "Diagnosis")
    display_age_corr(norm_data, "age_at_visit")
  } else if (dataset == "AIBL") {
    display_confounding(norm_data, "age_split", "Age Tertiles")
    display_confounding(norm_data, "sex", "Sex")
    display_confounding(norm_data, "diagnosis", "Diagnosis")
    display_age_corr(norm_data, "pred_age_blup")
  } else if (dataset == "FHS") {
    display_confounding(norm_data, "age_split", "Age Tertiles")
    display_confounding(norm_data, "sex", "Sex")
    display_confounding(norm_data, "AD_STATUS", "Diagnosis")
    display_age_corr(norm_data, "Age_at_collection")
  } else if (dataset == "MESA") {
    display_confounding(norm_data, "SEX", "Sex")
    display_confounding(norm_data, "race", "Race")
    display_age_corr(norm_data, "AGE_AT_COLLECTION")
  }
}
```

# Analyze

## ADNI Data

### Check Confounding

```{r}
norm_data <- load_normalized_data("ADNI")
display_confounding(norm_data, "age_split_by_tertiles", "Age Tertiles")
display_confounding(norm_data, "sex", "Sex")
display_confounding(norm_data, "DX", "Diagnosis")
display_age_corr(norm_data, "age_at_visit")
```

### Run Analysis

```{r}
plot_ls <- analyze_dataset("ADNI", expt_var = "age_split_by_tertiles")
```

### Plot PCAs

```{r, fig.width = 18, fig.height = 9}
gridExtra::grid.arrange(
  plot_ls$pca1$p1, plot_ls$pca1$p2, ncol = 2, top = "PCA by Batch"
)
gridExtra::grid.arrange(
  plot_ls$pca2$p1, plot_ls$pca2$p2, ncol = 2, top = "PCA by Age"
)
gridExtra::grid.arrange(
  plot_ls$pca3$p1, plot_ls$pca3$p2, ncol = 2, top = "PCA to Check Structure"
)
```

### KW Plots

```{r, fig.width = 18, fig.height = 30}
ggpubr::ggarrange(plotlist = plot_ls$kw_ls, ncol = 2, nrow = 3)
```

## AIBL Data

### Check Confounding

```{r}
norm_data <- load_normalized_data("AIBL")
display_confounding(norm_data, "age_split_by_tertiles", "Age Tertiles")
display_confounding(norm_data, "sex", "Sex")
display_confounding(norm_data, "diagnosis", "Diagnosis")
display_age_corr(norm_data, "pred_age_elastic")
```

### Run Analysis

```{r}
plot_ls <- analyze_dataset("AIBL", expt_var = "age_split_by_tertiles")
```

### Plot PCAs

```{r, fig.width = 18, fig.height = 9}
gridExtra::grid.arrange(
  plot_ls$pca1$p1, plot_ls$pca1$p2, ncol = 2, top = "PCA by Batch"
)
gridExtra::grid.arrange(
  plot_ls$pca2$p1, plot_ls$pca2$p2, ncol = 2, top = "PCA by Age"
)
gridExtra::grid.arrange(
  plot_ls$pca3$p1, plot_ls$pca3$p2, ncol = 2, top = "PCA to Check Structure"
)
```

### KW Plots

```{r, fig.width = 18, fig.height = 30}
ggpubr::ggarrange(plotlist = plot_ls$kw_ls, ncol = 2, nrow = 3)
```

## Framingham Data

### Check Confounding

```{r}
norm_data <- load_normalized_data("FHS")
display_confounding(norm_data, "age_split_by_tertiles", "Age Tertiles")
display_confounding(norm_data, "sex", "Sex")
display_confounding(norm_data, "AD_STATUS", "Diagnosis")
display_age_corr(norm_data, "Age_at_collection")
```

### Run Analysis

```{r}
plot_ls <- analyze_dataset("FHS", expt_var = "age_split_by_tertiles")
```

### Plot PCAs

```{r, fig.width = 18, fig.height = 9}
gridExtra::grid.arrange(
  plot_ls$pca1$p1, plot_ls$pca1$p2, ncol = 2, top = "PCA by Batch"
)
gridExtra::grid.arrange(
  plot_ls$pca2$p1, plot_ls$pca2$p2, ncol = 2, top = "PCA by Age"
)
gridExtra::grid.arrange(
  plot_ls$pca3$p1, plot_ls$pca3$p2, ncol = 2, top = "PCA to Check Structure"
)
```

### KW Plots

```{r, fig.width = 18, fig.height = 30}
ggpubr::ggarrange(plotlist = plot_ls$kw_ls, ncol = 2, nrow = 3)
```

## MESA Data

### Check Confounding

```{r}
norm_data <- load_normalized_data("MESA")
display_confounding(norm_data, "SEX", "Sex")
display_confounding(norm_data, "race", "Race")
display_age_corr(norm_data, "AGE_AT_COLLECTION")
```

### Run Analysis

```{r}
plot_ls <- analyze_dataset("MESA", expt_var = "race")
```

### Plot PCAs

```{r, fig.width = 18, fig.height = 9}
gridExtra::grid.arrange(
  plot_ls$pca1$p1, plot_ls$pca1$p2, ncol = 2, top = "PCA by Batch"
)
gridExtra::grid.arrange(
  plot_ls$pca2$p1, plot_ls$pca2$p2, ncol = 2, top = "PCA by Race"
)
gridExtra::grid.arrange(
  plot_ls$pca3$p1, plot_ls$pca3$p2, ncol = 2, top = "PCA to Check Structure"
)
```

### KW Plots

```{r, fig.width = 18, fig.height = 30}
ggpubr::ggarrange(plotlist = plot_ls$kw_ls, ncol = 2, nrow = 3)
```


# Session Information

<details>
  <summary>**Session Info**</summary>
```{r session}
sessionInfo()
```
</details>
